Combining information from photometric and spectroscopic data:\\ Probabilistic spectrophotometric inference

### 2015-05-15 tex/ms.tex
code and text for inference without logarithmic transform of the data

If you don't believe the reported data noise variances $\sigma_n^2$,
it often makes sense to add in quadrature a ``jitter'' variance $s^2$.
For photometric data this ``jitter'' is often best described as a
fractional error, $s^2 \equiv \hat{s}^2\mu(\theta)^2$, corresponding
for example to an additional magnitude zero-point uncertainty.  This
is a simple modification of the chi-squared log-likelihood:
\begin{eqnarray}\label{eq:photometricLF}
\ln p(d\given\theta,s^2) &=& -\frac{1}{2}\,\sum_{n=1}^N
                             \left[\frac{[d_n - \mu_n(\theta)]^2}
                             {[\sigma_n^2 + s^2]} + 
                             \ln(2\pi\,[\sigma_n^2 + s^2]) \right]
\quad .
\end{eqnarray}
This modification shows the necessity of including the log-variance
term.
This---equation~(\ref{eq:photometricLF})---is in fact the
log-likelihood function we are going to use for the photometric
measurements in what follows.  In this formulation, if $\theta$ are
the parameters of interest, the jitter variance $\hat{s} = 
s/\mu(\theta)$ is a nuisance noise-model parameter representing
additonal fractional uncertainty in the flux.  The Gaussian Process
will be a generalization of this modification.

For the spectroscopic measurements we are going to account for
the spectrophotometric calibration in our likelihood function.  We
take this calibration-vector that modifies the mean spectroscopic
model to produce the observed spectrum to be multiplicative with
respect to the mean model.  In this case we define the residual of the
data from the model in terms of the logarithmic differences.  This
logarithmic formulation requires that $d$ be everywhere positive, and
distorts the uncertainties, especially at low S/N ratio.  The
advantages of defining the residual in this way are that the model
remains linear (and therefore fast) and the calibration vector is
forced to be positive everywhere.  Additionally, at high signal to
noise the distortion of the uncertainties will be minimal.

{\color{blue} BDJ This paragraph should explain why we are using the
GP, not just how it works.}
The multiplicative calibration vector can be explicitly modeled, for
example using polynomials.  However, for complex calibration vectors
over large wavelength ranges, the number of parameters required to
model the calibration can become quite large (e.g. very high order
polynomials) and the correct parameterization may be unknown.  It is
therefore desirable to introduce a non-parameteric model of the
calibration.  We do this using a Gaussian Process: BDJ: prior over
functions, analytically marginalizable, covariant fractional
uncertainties the same as multiplying by a function (covariant
uncertainties the same as adding a function)

The calibration-vector and Gaussian-Process modified log-likelihood is
\begin{eqnarray}\label{eq:spectroscopicLF}
\ln p(d\given\theta,\alpha) &=&
                                -\frac{1}{2}\,\left[\transpose{\Delta}\,C^{-1}\,\Delta
                                + \ln([2\pi]^N\,\det{C}) \right]
\\
\Delta &\equiv& \ln d - \ln \mu(\theta) - f(\alpha) 
\\
C_{nn'} &\equiv& [\hat{\sigma}_n^2 + s^2]\,\delta_{nn'} +
K_\alpha(\lambda_n - \lambda_{n'})
\\
\hat{\sigma_n} & = &\sigma_n/d
\quad ,
\end{eqnarray}
where $\alpha$ is a blob of nuisance parameters
(which determine the jitter, the calibration vector and the kernel function),
$\Delta$ is a residual column vector,
$f(\alpha)$ is a calibration column vector,
$C$ is the Gaussian Process covariance matrix,
made up of $N^2$ values $C_{nn'}$,
$K_\alpha$ is a kernel function that sets the wavelength scale and
amplitude of the residual calibration issues,
$\hat{\sigma}_n$ are the fractional uncertainties of the data,
and each $\lambda_n$ is the wavelength of spectral pixel $n$.
The calibration vector $f(\alpha)$ is going to take care of the
large-scale calibration residual; the Gaussian Process is going to
pick up smaller-scale residuals away from \emph{that}.
The flexibility of the Gaussian Process specified by $C$ will obviate
the need for numerous high order polynomial terms in $f(\alpha)$ and
thus reduce the number of free parameters required in the model.  The
prior over functions implied by $C$ is analytically conditioned on the
residuals $\Delta$.
The residual column vector is defined in terms of the logarithmic
difference between the data and the mean model $\mu$ since we are
taking the calibration vector to be multiplicative with respect to the
mean model $\mu$. In this formulation the the mean model $\mu$ will be
effectively multiplied by $e ^{f(\alpha)}$
This---equation~(\ref{eq:spectroscopicLF})---is the log-likelihood we
are going to use for the spectroscopic data in what follows.

In detail, we are going to choose a $M$th-order polynomial form for
the calibration vector $f(\alpha)$ and a bell shape (Gaussian) for the
kernel function $K_\alpha(\cdot)$:
\begin{eqnarray}\displaystyle
f(\alpha) &=& c_0\,[1 + \sum_{m=1}^M c_m\,x^m]
\\
x &\equiv& \frac{\lambda}{\lambda_0} - 1
\\
K_\alpha(\lambda_n - \lambda_{n'}) &=& a^2\,\exp(-\frac{[\lambda_n - \lambda_{n'}]^2}{2\,\ell^2})
\\
\transpose{\alpha} &\equiv& \left[ \{c_m\}_{m=0}^M, s^2, a^2, \ell^2 \right]
\quad ,
\end{eqnarray}
where $\lambda$ is a column vector made up of the wavelengths $\lambda_m$
and $x$ is a scaled and shifted version of the same,
scaled by a fiducial wavelength $\lambda_0$.
Nuisance parameters $c_m$ set the shape of the large-scale calibration
issues, while $a^2$ and $\ell^2$ set the variance and correlation
length of the remaining covariant calibration noise.

XXX Alternative XXX
To avoid making the logarithmic transform described above, which
distorts the errors, we can formulate the problem in a different way,
modifying Equation \label{eq:spectroscopicLF} above as 
\begin{eqnarray}\label{eq:spectroscopicLF_v2}
\ln p(d\given\theta,\alpha) &=&
                                -\frac{1}{2}\,\left[\transpose{\Delta}\,C^{-1}\,\Delta
                                + \ln([2\pi]^N\,\det{C}) \right]
\\
\Delta &\equiv& d - \mu(\theta) * f(\alpha) 
\\
C_{nn'} &\equiv& [\sigma_n^2 + (\hat{s} * \mu)^2]\,\delta_{nn'} +
 K_\alpha(\lambda_n - \lambda_{n'}) \ast \mu \, \transpose{\mu}
\quad ,
\end{eqnarray}
where $\ast$ denotes element-wise multiplication and the
multiplication of $s$ and $K_\alpha$ by $\mu$ and
$\mu\,\transpose{\mu}$ respectively in the construction of $C_{nn'}$
means that they are defined in \emph{fractional} terms, and the
uncertainties remain Gaussian in linear flux space.  It remains to be
seen whether this actually works.

>>>>

If you don't believe the reported data noise variances $\sigma_n^2$,
it often makes sense to add in quadrature a ``jitter'' variance $s^2$.
This is a simple modification of the chi-squared log-likelihood:
\begin{eqnarray}\label{eq:photometricLF}
\ln p(d\given\theta,s^2) &=& -\frac{1}{2}\,\sum_{n=1}^N
                             \left[\frac{[d_n - \mu_n(\theta)]^2}
                             {[\sigma_n^2 + s^2]} + 
                             \ln(2\pi\,[\sigma_n^2 + s^2]) \right]
\quad .
\end{eqnarray}
This modification shows the necessity of including the log-variance
term.
This---equation~(\ref{eq:photometricLF})---is in fact the
log-likelihood function we are going to use for the photometric
measurements in what follows.  For photometric data this ``jitter'' is
often best described as a fractional error, $s^2 \equiv
\hat{s}^2\mu(\theta)^2$, corresponding for example to an additional
magnitude zero-point uncertainty.  In this formulation, if $\theta$
are the parameters of interest, the jitter variance $\hat{s} =
s/\mu(\theta)$ is a nuisance noise-model parameter representing
additonal fractional uncertainty in the flux.  The Gaussian Process
will be a generalization of this modification.

{\color{blue} BDJ This paragraph should explain why we are using the
GP, not just how it works.}
For the spectroscopic measurements we are going to account for the
spectrophotometric calibration in our likelihood function.  We take
this calibration-vector that modifies the mean spectroscopic model to
produce the observed spectrum to be multiplicative with respect to the
mean model.  The multiplicative calibration vector can be explicitly
modeled, for example using polynomials.  However, for complex
calibration vectors over large wavelength ranges, the number of
parameters required to model the calibration can become quite large
(e.g. very high order polynomials) and the correct parameterization
may be unknown.  It is therefore desirable to introduce a
non-parameteric model of the calibration.  We do this using a Gaussian
Process: BDJ: prior over functions, analytically marginalizable,
covariant fractional uncertainties the same as multiplying by a
function (covariant uncertainties the same as adding a function.)

The calibration-vector and Gaussian-Process modified log-likelihood is
\begin{eqnarray}\label{eq:spectroscopicLF_v2}
\ln p(d\given\theta,\alpha) &=&
                                -\frac{1}{2}\,\left[\transpose{\Delta}\,C^{-1}\,\Delta
                                + \ln([2\pi]^N\,\det{C}) \right]
\\
\Delta &\equiv& d - \mu(\theta) \ast f(\alpha) 
\\
C_{nn'} &\equiv& [\sigma_n^2 + (\hat{s} \, \mu)^2]\,\delta_{nn'} +
 K_\alpha(\lambda_n - \lambda_{n'}) \ast \mu \, \transpose{\mu}
\quad ,
\end{eqnarray}
where $\ast$ denotes element-wise multiplication,
$\theta$ are the physical parameters,
$\alpha$ is a blob of nuisance parameters
(which determine the jitter, the calibration vector and the kernel function),
$\Delta$ is a residual column vector,
$f(\alpha)$ is a calibration column vector,
$C$ is the Gaussian Process covariance matrix,
made up of $N^2$ values $C_{nn'}$,
$K_\alpha$ is a kernel function that sets the wavelength scale and
amplitude of the residual calibration issues,
and each $\lambda_n$ is the wavelength of spectral pixel $n$.
The multiplication of $\hat{s}$ and $K_\alpha$ by $\mu$ and
$\mu\,\transpose{\mu}$ respectively in the construction of $C_{nn'}$
signifies that they are defined in \emph{fractional} terms.
The calibration vector $f(\alpha)$ is going to take care of the
large-scale calibration residual; the Gaussian Process is going to
pick up smaller-scale residuals away from \emph{that}.
The flexibility of the Gaussian Process specified by $C$ will obviate
the need for numerous high order polynomial terms in $f(\alpha)$ and
thus reduce the number of free parameters required in the model.  The
prior over functions implied by $C$ is analytically conditioned on the
residuals $\Delta$.
This---equation~(\ref{eq:spectroscopicLF})---is the log-likelihood we
are going to use for the spectroscopic data in what follows.

In detail, we are going to choose the exponential of a $M$th-order
polynomial form for the calibration vector $f(\alpha)$ and a bell
shape (Gaussian) for the kernel function $K_\alpha(\cdot)$:
\begin{eqnarray}\displaystyle
f(\alpha) &=& \exp\left(\sum_{m=o}^M c_m\,x^m\right)
\\
x &\equiv& \frac{\lambda}{\lambda_0} - 1
\\
K_\alpha(\lambda_n - \lambda_{n'}) &=& a^2\,\exp\left(-\frac{[\lambda_n - \lambda_{n'}]^2}{2\,\ell^2}\right)
\\
\transpose{\alpha} &\equiv& \left[ \{c_m\}_{m=0}^M, s^2, a^2, \ell^2 \right]
\quad ,
\end{eqnarray}
where $\lambda$ is a column vector made up of the wavelengths
$\lambda_m$ and $x$ is a scaled and shifted version of the same,
scaled by a fiducial wavelength $\lambda_0$.  The exponentiation of
the polynomial in ensures that $f(\alpha)$ is everywhere positive.
Nuisance parameters $c_m$ set the shape of the large-scale calibration
issues, while $a^2$ and $\ell^2$ set the variance and correlation
length of the remaining covariant calibration noise.

XXX logarithmic Alternative XXX
In this case we define the residual of the data from the model in
terms of the logarithmic differences.  This logarithmic formulation
requires that $d$ be everywhere positive, and distorts the
uncertainties, especially at low S/N ratio.  The advantages of
defining the residual in this way are that the model remains linear
(and therefore fast) and the calibration vector is forced to be
positive everywhere.  Additionally, at high signal to noise the
distortion of the uncertainties will be minimal.

The calibration-vector and Gaussian-Process modified log-likelihood is
\begin{eqnarray}\label{eq:spectroscopicLF}
\ln p(d\given\theta,\alpha) &=&
                                -\frac{1}{2}\,\left[\transpose{\Delta}\,C^{-1}\,\Delta
                                + \ln([2\pi]^N\,\det{C}) \right]
\\
\Delta &\equiv& \ln d - \ln \mu(\theta) - \ln f(\alpha) 
\\
C_{nn'} &\equiv& [\hat{\sigma}_n^2 + s^2]\,\delta_{nn'} +
K_\alpha(\lambda_n - \lambda_{n'})
\\
\hat{\sigma_n} & = &\sigma_n/d
\quad ,
\end{eqnarray}
where $\hat{\sigma}_n$ are the fractional uncertainties of the data.
The residual column vector is defined in terms of the logarithmic
difference between the data and the mean model $\mu$ since we are
taking the calibration vector to be multiplicative with respect to the
mean model $\mu$. In this formulation the the mean model $\mu$ will be
effectively multiplied by $e ^{f(\alpha)}$
